Online Sequential Monte Carlo smoother for partially observed diffusion processes

### 2017-09-13 TexSubmission/ParisEM.tex
small changes to precise the necessity of a positive and unbiased estimator. Large changes in the SINE model section to precise the trade off for the choose of A1 bound or A3 bound. Plus answer to reviewers. Two figures to illustrate the choice of bound trade off

\begin{abstract}
This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed stochastic differential equations. 
This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e. as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples. The original algorithm cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown. 
We prove that it may be extended to partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators. 
This estimator is consistent and its performance are illustrated using data from two models. 
\end{abstract}

@@

In this paper, we propose to use Sequential Monte Carlo (SMC) methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending PaRIS algorithm.
The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
 The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. 
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. 
 The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother. 
 The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a positive and unbiased estimator.

@@

The algorithm proposed in this paper is based on Sequential Monte Carlo methods which offer a flexible framework to approximate such distributions with weighted empirical measures associated with random samples. At each time step, the samples are moved randomly in $\mathbb{R}^d$ and  associated with importance weights. In general situations, the computation of these importance weights involve the unknown transition density of the process \eqref{eq:target:sde}. The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities. Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of transition densities. Those estimators require different assumptions on the model \eqref{eq:target:sde}, we provide below several solutions that can be investigated.

\paragraph{General Poisson estimators}
This paper focuses mainly on GPEs which have been widely used recently and applied in a variety of disciplines. These estimators require that the diffusion coefficient $\Gamma$ is constant and equal to the identity matrix, see \cite{fearnhead:papaspiliopoulos:roberts:2008}. They may be applied to reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
\begin{equation}
\label{eq:diff:lamperti}
\mathrm{d}Z_t = \beta(Z_t)\mathrm{d}t + \mathrm{d}W_t\eqsp.
\end{equation}
By Ito's formula,  it is straightforward to show that, in the case of a reducible diffusion, the Jacobian matrix of $\eta$ satisfies
\[
\nabla \eta = \Gamma^{-1}\eqsp,
\]
and, in the case $d=1$,
\[
\beta:u \mapsto \frac{\alpha(\eta^{-1}(u))}{\Gamma(\eta^{-1}(u))} - \frac{\Gamma'(\eta^{-1}(u))}{2}\eqsp.	
\]
In the case of a scalar diffusion, this Lamperti transform is given by:
\[
\eta:x\mapsto \int_{x_0}^x\Gamma^{-1}(u)\mathrm{d}u\eqsp.
\]
In the general case, \cite[Proposition~$1$]{ait-sahalia:2008} shows that when $\Gamma$ is nonsingular, the SDE is reducible if and only if, for all $1\le i,j,k\le d$,
\begin{equation}
\label{eq:reducibility}
\frac{\partial\Gamma^{-1}_{i,j}}{\partial x_k} = \frac{\partial\Gamma^{-1}_{i,k}}{\partial x_j}\eqsp.
\end{equation}
In the case of a diagonal matrix $\Gamma$ \eqref{eq:reducibility} is equivalent to assume that $\Gamma$ is such that for each $1\le i\le d$, $\Gamma_{i,i}$ depends on $x_i$ only. \cite{ait-sahalia:2008} notes that the reducibility condition \eqref{eq:reducibility} holds also for some nondiagonal matrices $\Gamma$.
 This is true in particular in the case $d=2$ for stochastic volatility models where $\sigma$ is of the form:
\[
\Gamma(x) = \begin{pmatrix}a(x_1) & a(x_1)b(x_2)\\0&c(x_2)\end{pmatrix}\eqsp.
\]
GPEs consider that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e. we consider a diffusion after the application of the Lamperti transform.
 In addition, designing GPEs also requires that:
\begin{enumerate}[i)]
\item $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function ;
\item the function $x\mapsto \left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2$ is lower bounded where $\triangle$ is the Laplace operator.
\end{enumerate}
Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however it has natural applications in many fields such as movement ecology, see \cite{gloaguen:etienne:lecorff:2017}. Assumption ii) is a technical condition which ensures that exact sampling of processes solution to  \eqref{eq:target:sde} using acceptance rejection methods, see for instance \cite{beskos:papaspiliopoulos:roberts:2006,beskos:papaspiliopoulos:roberts:2008,fearnhead:papaspiliopoulos:roberts:2008}.

\paragraph{Continuous importance sampling based estimators}
In the case the previous assumptions are not fulfilled, alternatives to GPEs are given by continuous importance sampling procedures for SDE. In \cite{wagner:1989}, for each $0\le k \le n-1$, the transition density between $t_k$ and $t_{k+1}$ is expressed as an infinite expansion obtained using the Kolmogorov backward operator associated with  \eqref{eq:target:sde}. This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times using tractable proposal distributions (for instance based on an Euler discretization of the original SDE). Then, these samples are associated with random weights to ensure that the proposed estimator is unbiased. More recently, \cite{fearnhead:latuszynski:roberts:sermaidis:2017} extended the discrete time importance sampling estimator by introducing updates at random times associated with a renewal process. The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients and are therefore tractable.

The unbiasedness of these procedures as long as the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}. Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.

\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

@@

\paragraph{Bounded estimator of $q_k$ using GPEs}
For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
\begin{align*}
q_k(x,y)=\varphi_{\Delta_k}(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta_k}}\left[ \exp \left\lbrace - \int_0^{\Delta_k} \phi(\mw_s)\rmd s \right\rbrace \right]\eqsp,
\end{align*}
where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f. of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated at $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
\[
\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp.
\]
Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp$. 
As shown in \cite{fearnhead:papaspiliopoulos:roberts:2008}, a positive unbiased estimator is given by 
\begin{multline}
\widehat{q}_k(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta_k^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
\end{multline}
Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, where $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$. 
In that case, the estimator \eqref{eq:unbiased:q} becomes:
\begin{equation}
\widehat{q}_{k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
\end{equation}
On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to satisfy of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
\begin{align}
\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\label{eq:rho:func}
\end{align}
is upper bounded almost surely by $\hat{\sigma}^k_+$.
In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used. 
This condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.

@@

Assumption H\ref{assum:boundmodel} only involves the marginal likelihood $g_k$ of the observations and does not depend on the unbiased estimation of the transition density. In the case where the observations are given as in the Section~\ref{sec:exp}, this assumption holds as soon as the variance of the observation is bounded away fro zero.

\begin{hypA}
\label{assum:boundalgo}
$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widehat{\omega}_{k}|_{\infty} < \infty$, where
\[
\widehat{\omega}_{0}(x) = \frac{\chi(x)g_0(x)}{\eta_0(x)} \quad\mbox{and for}\; k\ge1\quad\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{q}_{k-1}(x,x';z)g_{k}(x')}{\vartheta_{k}(x) p_{k-1} (x,x')}\eqsp.
\]
\end{hypA}
Assumption H2 depends on the algorithm used to estimate the transition densities and on the tuning parameters of the  SMC filter. The most common choice is $\vartheta_k = 1$ so that under H1, the only requirement is to control $\widehat{q}_{k-1}$ and $p_{k-1}$. For instance, in the case of the GPE-1, as explained in Section~3, H2 is satisfied if $\phi$ is upper bounded (as for the EA1).

@@

In order to evaluate the performance of the proposed algorithm, the following strategy has been chosen. 
We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50). 
The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model and Figure~\ref{fig:res:LG:EM} for the log-growth model. 
This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.  
Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
In order to obtain a reference value for our study, the intermediate quantity of the EM algorithm is also estimated 30 times using the GRand PaRIS algorithm with $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$. 
Figures \ref{fig:res:SINE:EM} and \ref{fig:res:LG:EM} display this estimate for an example with one simulated data set. 
The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\mu\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
\end{equation}
This simple model has no explicit transition density, however GPEs may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
\begin{equation*}
Y_k = X_k + \varepsilon_k\eqsp,
\end{equation*}
where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0, \sigma^2_\text{obs})$, the resulting set of model parameters is $\theta = (\mu, \sigma_\text{obs})$.
In the example displayed in Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ and $ \sigma_\text{obs} = 1$.
In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.

@@

\bibitem{gloaguen:etienne:lecorff:2017}
P.~ Gloaguen, M.-P.~\'Etienne, and S.~Le~Corff.
\newblock Inference procedures to estimate potential based movement models in ecology.
\newblock {\em To appear in the {J}ournal of the {R}oyal {S}tatistical {S}ociety: {S}eries {C}}.

>>>>

\begin{abstract}
This paper introduces a new algorithm to approximate smoothed additive functionals for partially diffusion processes. 
This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e. as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples. The original algorithm cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown. 
We prove that it may be extended to partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators. 
This estimator is consistent and its performance are illustrated using data from two models. 
\end{abstract}

@@

In this paper, we propose to use SMC methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending the PaRIS algorithm.
The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
 The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of the PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. 
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. 
 The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother. 
 The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a positive and unbiased estimator.

@@

The algorithm proposed in this paper is based on Sequential Monte Carlo methods which offer a flexible framework to approximate such distributions with weighted empirical measures associated with random samples. 
At each time step, the samples are moved randomly in $\mathbb{R}^d$ and  associated with importance weights. 
In general situations, the computation of these importance weights involve the unknown transition density of the process \eqref{eq:target:sde}. 
The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities. Moreover, this estimator must be almost surely positive and upper bounded.
Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of transition densities. 
Those estimators require different assumptions on the model \eqref{eq:target:sde}, we provide below several solutions that can be investigated.

\paragraph{General Poisson estimators}
This paper focuses mainly on GPEs which have been widely used recently and applied in a variety of disciplines. These estimators require that the diffusion coefficient $\Gamma$ is constant and equal to the identity matrix, see \cite{fearnhead:papaspiliopoulos:roberts:2008}. They may be applied to reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
\begin{equation}
\label{eq:diff:lamperti}
\mathrm{d}Z_t = \beta(Z_t)\mathrm{d}t + \mathrm{d}W_t\eqsp.
\end{equation}
By Ito's formula,  it is straightforward to show that, in the case of a reducible diffusion, the Jacobian matrix of $\eta$ satisfies
\[
\nabla \eta = \Gamma^{-1}\eqsp,
\]
and, in the case $d=1$,
\[
\beta:u \mapsto \frac{\alpha(\eta^{-1}(u))}{\Gamma(\eta^{-1}(u))} - \frac{\Gamma'(\eta^{-1}(u))}{2}\eqsp.	
\]
In the case of a scalar diffusion, this Lamperti transform is given by:
\[
\eta:x\mapsto \int_{x_0}^x\Gamma^{-1}(u)\mathrm{d}u\eqsp.
\]
In the general case, \cite[Proposition~$1$]{ait-sahalia:2008} shows that when $\Gamma$ is nonsingular, the SDE is reducible if and only if, for all $1\le i,j,k\le d$,
\begin{equation}
\label{eq:reducibility}
\frac{\partial\Gamma^{-1}_{i,j}}{\partial x_k} = \frac{\partial\Gamma^{-1}_{i,k}}{\partial x_j}\eqsp.
\end{equation}
In the case of a diagonal matrix $\Gamma$ \eqref{eq:reducibility} is equivalent to assume that $\Gamma$ is such that for each $1\le i\le d$, $\Gamma_{i,i}$ depends on $x_i$ only. \cite{ait-sahalia:2008} notes that the reducibility condition \eqref{eq:reducibility} holds also for some nondiagonal matrices $\Gamma$.
 This is true in particular in the case $d=2$ for stochastic volatility models where $\sigma$ is of the form:
\[
\Gamma(x) = \begin{pmatrix}a(x_1) & a(x_1)b(x_2)\\0&c(x_2)\end{pmatrix}\eqsp.
\]
GPEs consider that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e. we consider a diffusion after the application of the Lamperti transform.
 In addition, designing GPEs also requires that:
\begin{enumerate}[i)]
\item $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function ;
\item the function $x\mapsto \left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2$ is lower bounded where $\triangle$ is the Laplace operator.
\end{enumerate}
Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however it has natural applications in many fields such as movement ecology, see \cite{gloaguen:etienne:lecorff:2017}. 
Assumption ii) is a technical condition which ensures that exact sampling of processes solution to  \eqref{eq:target:sde} using acceptance rejection methods, see for instance \cite{beskos:papaspiliopoulos:roberts:2006,beskos:papaspiliopoulos:roberts:2008,fearnhead:papaspiliopoulos:roberts:2008}.
In addition to provide an unbiased estimate of the transition density, the GPE ensure that this estimate is almost surely positive. 
Moreover, as detailled below, one can define, under certain conditions, a GPE that is almost surely upper bounded.
\paragraph{Continuous importance sampling based estimators}
In the case the previous assumptions are not fulfilled, alternatives to GPEs are given by continuous importance sampling procedures for SDE. 
In \cite{wagner:1989}, for each $0\le k \le n-1$, the transition density between $t_k$ and $t_{k+1}$ is expressed as an infinite expansion obtained using the Kolmogorov backward operator associated with  \eqref{eq:target:sde}. 
This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times using tractable proposal distributions (for instance based on an Euler discretization of the original SDE). 
Then, these samples are associated with random weights to ensure that the proposed estimator is unbiased. More recently, \cite{fearnhead:latuszynski:roberts:sermaidis:2017} extended the discrete time importance sampling estimator by introducing updates at random times associated with a renewal process. 
The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients and are therefore tractable.

The unbiasedness of these procedures as long as the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}. 
Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.
In addition to unbiasedness, the proposed algorithm in this work requires that the estimator of the transition density is almost surely positive and upper bounded. 
It is beyond the scope of this paper to define a continuous importance sampling estimator with such properties, but this could be an interesting perspective if it requires less assumptions than the GPE.
\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

@@

\paragraph{Bounded estimator of $q_k$ using GPEs}
For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
\begin{align*}
q_k(x,y)=\varphi_{\Delta_k}(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta_k}}\left[ \exp \left\lbrace - \int_0^{\Delta_k} \phi(\mw_s)\rmd s \right\rbrace \right]\eqsp,
\end{align*}
where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f. of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated at $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
\[
\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp.
\]
Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp$. 
As shown in \cite{fearnhead:papaspiliopoulos:roberts:2008}, a positive unbiased estimator is given by 
\begin{multline}
\widehat{q}_k(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta_k^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
\end{multline}
Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, where $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$. 
In that case, the estimator \eqref{eq:unbiased:q} becomes:
\begin{equation}
\widehat{q}_{k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
\end{equation}
On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1.
 Therefore, a sufficient condition to satisfy one of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
\begin{align*}
\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}
\end{align*}
is upper bounded almost surely by $\hat{\sigma}^k_+$.
In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:GPE1} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used. 
This condition is always satisfied for models in the domains required for the applications of  exact algorithms EA1, EA2 and EA3 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.

@@

Assumption {\bf H\ref{assum:boundmodel}} only involves the marginal likelihood $g_k$ of the observations and does not depend on the unbiased estimation of the transition density. 
In the case where the observations are given as in the Section~\ref{sec:exp}, this assumption holds as soon as the variance of the observation is bounded away from zero.

\begin{hypA}
\label{assum:boundalgo}
$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widehat{\omega}_{k}|_{\infty} < \infty$, where
\[
\widehat{\omega}_{0}(x) = \frac{\chi(x)g_0(x)}{\eta_0(x)} \quad\mbox{and for}\; k\ge1\quad\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{q}_{k-1}(x,x';z)g_{k}(x')}{\vartheta_{k}(x) p_{k-1} (x,x')}\eqsp.
\]
\end{hypA}
Assumption {\bf H\ref{assum:boundalgo}} depends on the algorithm used to estimate the transition densities and on the tuning parameters of the  SMC filter. 
The most common choice is $\vartheta_k = 1$ so that under H1, the only requirement is to control $\widehat{q}_{k-1}$ and $p_{k-1}$. 
For instance, in the case of the GPE-1, as explained in Section~3, {\bf H\ref{assum:boundalgo}} is satisfied if $\phi$ is upper bounded (as for the EA1).

@@

In order to evaluate the performance of the proposed algorithm, the following strategy has been chosen. 
We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50). 
The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model and Figure~\ref{fig:res:LG:EM} for the log-growth model. 
This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.  
Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
In order to obtain a reference value for our study, the intermediate quantity of the EM algorithm is also estimated 30 times using the GRand PaRIS algorithm with $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$. 
Figures \ref{fig:res:SINE:EM} and \ref{fig:res:LG:EM} display this estimate for an example with one simulated data set. 
The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times, resulting a fair comparison. On a personnal computer \footnote{(i7-6600U CPU @ 2.60GHz)}, for the parameters mentionned above, it takes around 25 seconds per E step to be performed.

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\mu\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
\end{equation}
This simple model has no explicit transition density, however GPEs may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
\begin{equation*}
Y_k = X_k + \varepsilon_k\eqsp,
\end{equation*}
where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0, \sigma^2_\text{obs})$, the resulting set of model parameters is $\theta = (\mu, \sigma_\text{obs})$.
In the example displayed in Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ and $ \sigma_\text{obs} = 1$.

In the case of the SINE model, the estimator $\widehat{q}_k(\cdot)$ defined by equation \eqref{eq:GPE1} satisfies both \eqref{assumpt:global} and \eqref{assumpt:ysupport}. 
Therefore, the corresponding bound $\widehat{\sigma}_+^k$ can be obtained using numerical optimization. 
If that bound is chosen, the GRand PaRIS algorithm has linear complexity in the number of particles. 
As an alternative, it is worth noting here that the bound $\widehat{\sigma}_+^k$ defined by \eqref{assumpt:local} can also be used. This method has a quadratic cost in the number of particles, but provides the optimal bound for the algorithm of Lemma \ref{lem:AR:unbiased}, and therefore reduces the expected time before acceptance, specially when the time step $\Delta_k$ is large. 
Therefore, depending on the number of particles wanted and the time step $\Delta_k$, one has to choose whether to use the  $\widehat{\sigma}_+^k$ defined by \eqref{assumpt:global}, \eqref{assumpt:ysupport} or \eqref{assumpt:local}. In the configuration presented here, both bounds resulted in an equivalent computation time.

@@

\bibitem{gloaguen:etienne:lecorff:2017}
P.~ Gloaguen, M.-P.~\'Etienne, and S.~Le~Corff.
\newblock Stochastic differential equation based on a multimodal potential to model movement data in ecology\textit{•}.
\newblock {\em To appear in the {J}ournal of the {R}oyal {S}tatistical {S}ociety: {S}eries {C}}.
